---
title: "HISAT2-StringTie-Ballgown analysis"
author: "German Novakovskiy"
date: "June 18, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# RNA-seq analysis with Ballgown

```{r}
suppressMessages(suppressWarnings(library(AnnotationDbi)))
suppressMessages(suppressWarnings(library(org.Hs.eg.db)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(tximport)))
suppressMessages(suppressWarnings(library(ballgown)))
suppressMessages(suppressWarnings(library(tibble)))
suppressMessages(suppressWarnings(library(reshape2)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(RColorBrewer)))
suppressMessages(suppressWarnings(library(ermineR)))
suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(tximport)))
suppressMessages(suppressWarnings(library(readr)))
suppressMessages(suppressWarnings(library(tximportData)))
suppressMessages(suppressWarnings(library(GenomicFeatures)))
suppressMessages(suppressWarnings(library(biomaRt)))
suppressMessages(suppressWarnings(library(gplots)))
suppressMessages(suppressWarnings(library(geneplotter)))
suppressMessages(suppressWarnings(library(fdrtool)))
suppressMessages(suppressWarnings(library(genefilter)))
suppressMessages(suppressWarnings(library(topGO)))
suppressMessages(suppressWarnings(library(devtools)))
suppressMessages(suppressWarnings(library(RSkittleBrewer)))
```

## Running HISAT2 (genome alignment)

Remove all # for running:

```{bash}
##Downloading genome for hisat2
#curl ftp://ftp.ensembl.org/pub/grch37/release-92/fasta/homo_sapiens/dna/Homo_sapiens.GRCh37.dna_sm.primary_assembly.fa.gz -o Homo_sapiens.GRCh37.dna_sm.primary_assembly.fa.gz

##Downloading transcriptome annotation
#curl ftp://ftp.ensembl.org/pub/grch37/release-92/gtf/homo_sapiens/Homo_sapiens.GRCh37.87.gtf.gz -o genes_92_19.gtf.gz

##indexing genome with hisat2
#hisat2-build -p 8 Homo_sapiens.GRCh38.dna_sm.primary_assembly.fa Homo_sapiens.GRCh38.dna_sm.primary_assembly
```

Running HISAT2. Remove all # for running, except first line (when it's #!/bin/bash):

```{bash}
#!/bin/bash

#samples=($(ls good_reads))

#for i in "${samples[@]}"
#do

        #echo "Started with $i"

        ##all R1 reads for this sample from different libraries
        #TEST1=$(ls good_reads/"$i"/*R1* | tr '\n' ' ' | \
        #awk '{for(i=1;i<=NF;i++){printf $i;if(i<NF)printf ","}}')

        ##all R2 reads for this sample from different libraries
        #TEST2=$(ls good_reads/"$i"/*R2* | tr '\n' ' ' | \
        #awk '{for(i=1;i<=NF;i++){printf $i;if(i<NF)printf ","}}')

        ##running hisat2 alignment of cdna reads
        #run_rnacocktail.py align --align_idx Homo_sapiens.GRCh37.dna_sm.primary_assembly --outdir hisat2_output \
        #--workdir hisat2_work --ref_gtf genes_92_19.gtf --1 $TEST1 \
        #--2 $TEST2 --hisat2 ../hisat2-2.0.5/hisat2 \
        #--hisat2_sps ../hisat2-2.0.5/hisat2_extract_splice_sites.py --samtools ../samtools-1.8/samtools --sample $i

        #echo '\n'
        #echo '\n'
        #echo "Done with $i"
        #echo '\n'
        #echo '\n'
#done

```

Running StringTie. Remove all # for running, except first line (when it's #!/bin/bash):

```{bash}
#!/bin/bash

##Running StringTie for the first time
#samples=($(ls hisat2_output/hisat2))

#for i in "${samples[@]}"
#do

        #echo "Started with $i"

        #running stringtie with hisat2 alignment files
        #../stringtie-1.3.4d.Linux_x86_64/stringtie hisat2_output/hisat2/$i/alignments.sorted.bam \
        #-l $i -o stringtie_result/$i.gtf -G genes_92_19.gtf --rf -A actual_stringtie_test_out/$i.tab

        #echo "Done with $i"
        
#done

##Constructing merge list (what files to merge)
#ls -R stringtie_result | awk '
#/:$/&&f{s=$0;f=0}
#/:$/&&!f{sub(/:$/,"");s=$0;f=1;next}
#NF&&f{ print s"/"$0 }' > mergelist.txt

##merge gtf files for different samples
#../stringtie-1.3.4d.Linux_x86_64/stringtie --merge -G genes_92_19.gtf -o stringtie_result/stringtie_merged.gtf mergelist.txt

##run gffcompare to compare merged gtf file with reference gtf file (genes_92_19.gtf)
#../comparegff/gffcompare/gffcompare -r genes_92_19.gtf -o compare_res stringtie_result/stringtie_merged.gtf

##rerunning stringtie with merged gtf (-e is used, because now we want to quantifu only transcripts from
##the merged file)
##-B parameter is used in order to generate files for Ballgown
#for j in "${samples[@]}"
#do

        #echo "Started with $j"

        #RErunning stringtie with hisat2 alignment files and new merged gtf file
        #../stringtie-1.3.4d.Linux_x86_64/stringtie hisat2_output/hisat2/$j/alignments.sorted.bam \
        #-e -B -o stringtie_result_merged/$j/$j.gtf -G stringtie_result/stringtie_merged.gtf --rf

        #echo "Done with $j"

#done

##counting transcripts and genes for DESeq2 (45 is an average read length for this libraries)
#python prepDE.py -i prepde_input.txt -l 45 

```

Results of gffcompare:
```{bash}
less ~/RNA-seq_Cynthia/compare_res.stats
```

High sensitivity here means that all of the StringTie transcripts match the known transcripts (in genes_19_92.gtf annotation file). The precision is lower (but not much) indicating that many of the StringTie transcripts are not in the list of known transcripts, which are either false positives or truly de novo transcripts (we see that there are not a lot of them). The novel exons, introns, and loci indicate how many of the sites were not found in the list of known transcripts.

## Ballgown analysis of transcripts 

```{r}
#metadata construction
ids <- c("Ctl1-1", "Ctl1-2", "Ctl2-1", "Ctl2-2", "VAN-011-1", "VAN-011-2",
             "VAN-012-1", "VAN-012-2", "VAN-013-1", "VAN-013-2", "VAN0092", "VAN3360091")
#ballgown is not good in handling technical replicates, thus I will include
#an additional column here
samples <- c("Ctl1", "Ctl1", "Ctl2", "Ctl2", "VAN011", "VAN011",
             "VAN012", "VAN012", "VAN013", "VAN013", "VAN0092", "VAN0092")
sex <- c("M", "M", "M", "M", "F", "F", "F", "F", "M", "M", "F", "F")
national <- c("Caucasian", "Caucasian", "Asian", "Asian",
              "Caucasian", "Caucasian", "Caucasian", "Caucasian",
              "Caucasian", "Caucasian", "Caucasian", "Caucasian")
relation <- c("NoFamily", "NoFamily", "NoFamily", "NoFamily",
              "Family", "Family", "Family", "Family",
              "Family", "Family", "Family", "Family")

#1 is affected, 0 if not
affected <- c(0, 0, 0, 0, 1, 1, 0, 0, 1, 1, 1, 1)
rep <- c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6)
metadata <- data.frame(Ids = ids, Samples = samples, National = national,
                       Relation = relation, Sex = sex, Affected = affected, Replicate = rep)
metadata$Affected <- as.factor(metadata$Affected)
metadata$Samples <- as.factor(metadata$Samples)
metadata$Replicate <- as.factor(metadata$Replicate)
rownames(metadata) <- ids
metadata
```

Read in the expression data that were calculated by StringTie:

```{r}
dir <- "~/RNA-seq_Cynthia/ballgown"
files <- file.path(dir, metadata$Ids)
files
```

```{r}
bg <-  ballgown(dataDir = "~/RNA-seq_Cynthia/ballgown", 
                samples = files, 
                pData = metadata)
bg
```

Filter to remove low-abundance genes. Here we remove all transcripts with a variance across samples less than one:

```{r}
bg_filt = subset(bg,"rowVars(texpr(bg)) >1",genomesubset=TRUE)

bg_filt
```

Identify transcripts that show statistically significant differences between groups.Note that Ballgown’s statistical test is a standard linear model-based comparison. For small sample sizes (n < 4 per group), it is often better to perform regularization. 

```{r}
results_transcripts = stattest(bg_filt, feature="transcript",covariate="Affected",
                               adjustvars =c("Sex"), getFC=TRUE, meas="FPKM")
```

Add gene names and gene IDs to the results_transcripts data frame:

```{r}
results_transcripts <-  data.frame(geneNames=ballgown::geneNames(bg_filt),
                                   geneIDs=ballgown::geneIDs(bg_filt), results_transcripts)

```

Sort the results from the smallest P value to the largest:

```{r}
results_transcripts <- arrange(results_transcripts,qval)
head(results_transcripts, 10)
```

```{r}
plotTranscripts(ballgown::geneIDs(bg)[49757], bg, main=c('Gene OASL'), sample=ids)
```
```{r}
#3rd chromosome (start-end 82854807	82855108), 1 exon, probably FP
plotTranscripts(ballgown::geneIDs(bg)[143824], bg, main=c('Gene MSTRG.21534'), sample=ids)
```

```{r}
#17th chromosome (start-end 20461500	20461760), 1 exon, probably FP
plotTranscripts(ballgown::geneIDs(bg)[84373], bg, main=c('Gene MSTRG.12756'), sample=ids)
```
```{r}
plotTranscripts(ballgown::geneIDs(bg)[101577], bg, main=c('Gene FARSA'), sample=ids)
```
```{r}
#17 chromosome (gene name ZNHIT3)
plotTranscripts(ballgown::geneIDs(bg)[86159], bg, main=c('Gene MSTRG.13094'), sample=ids)
```

```{r}
#17 chromosome (gene name ZNHIT3)
plotTranscripts(ballgown::geneIDs(bg)[148450], bg, main=c('Gene RPL22L1'), sample=ids)
```


